---
title: "NFL Quarterback passing"
author: "Andrew Mellor"
date: "06/05/2020"
output:
  html_document:
    toc: true
    toc_depth: 3
---

```{r Setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

The purpose of this document is to answer these questions:

**How does passer rating relate to success?
We can use the number of wins (wins) a team has when a particular quarterback starts as a measure of success. And then the questions we are trying to answer now are:
Do quarterbacks with a higher passer rating have more wins?
Do quarterbacks with a lower passer rating have fewer wins?
How many wins can we expect to get when we have a starting quarterback with a particular passer rating?**

We can use simple linear regression to answer these questions.

## Load required packages

```{r Packages, message=FALSE}
#Include all required packages at the start
library(tidyverse) 
library(ggplot2)
library(broom)
```

## Data

The data we will be using for this activity can be found in the 2019_nfl_qb_data.csv file and contains Quarterback stats for all quarterbacks in the 2019 NFL season. It has been sourced from [pro-football-reference.com](https://www.pro-football-reference.com/years/2019/passing) and consists of the following variables:

The [data](https://unicanberra.instructure.com/courses/7493/files/1778304) was provided by Dr Jocelyn Mara from the University of Canberra^[University of Canberra. Applied Data Analysis in Sport PG (10157)].

**It is probably important to note some of the data points that are not self explanatory:**

1. case_no: unique observation identification number

2. player_name: quarterback player name

3. player_id: unique player identification number

4. team: team the quarterback plays for in the NFL

5. age: age of player in years

6. position: Uppercase QB indicates a primary starting quarterback, while lowercase qb

7. indicates a secondary quarterback

8. games_played: number of games played

9. games_started: number of games started

10. wins: number of games won when starting

11. losses: number of games lost when starting

12. draws: number of games drawn when starting

13. completions: number of passes completed

14. attempts: number of passes attempted

15. cmp_pc: passes completed as a percentage of passes attempted

16. yards: yards gained by passing

17. touchdowns: number of passing touchdowns

18. touchdown_pc: passing touchdowns as a percentage of passes attempted

19. interceptions: number of interceptions

20. interceptions_pc: interceptions as a percentage of passes attempted

21. passer_rating: Passer rating (also known as quarterback rating). For a description of the Passer rating see [here](https://en.wikipedia.org/wiki/Passer_rating)

22. sacks: number of [sacks](https://en.wikipedia.org/wiki/Quarterback_sack)

23. yards_lost: number of yards lost due to sacks

24. sack_pc: sacks as a percentage of passes attempted

25. Each row is one players total stats for the 2019 season.

## Read in Data

Read in the `2019_nfl_qb_data.csv` file using the `read_csv()` function from the `readr` package. 

```{r Read_Data, message=FALSE} 
#Read in the data
#data dir should be in the working directory
df <- read_csv("data/2019_nfl_qb_data.csv") 
```

## Checking the data
```{r Structure, message=FALSE, results='hide'}
str(df) #provides structure of df


head(df) #shows first 6 rows of df


tail(df) #shows last 6 rows of df
```


```{r Na, message=FALSE, results='hide'}
#Check for missing values
sum(is.na(df)) 
```

## Data Transformation
For this step, do the following in one sequence of code:

1. filter the data so only quarterbacks that have started a minimum of 10 games are included,
2. create a new variable called win_percentage, that expresses the number of wins as a percentage of games_started
Save the result in a new object.
````{r Data Transformation}
#Create win_percentage
df2 <- df %>%
filter(games_started >= 10) %>%
mutate(win_percentage = wins / games_started * 100)

view(df2)
````

## Exploratory analysis

### Relationship between passer_rating and win_percentage

```` {r Plot1, echo=FALSE}
#Plot the relationship between paser_rating (x-axis) & win_percentage (y-axis)
ggplot(df2, aes(x = passer_rating, y = win_percentage))+
geom_point(alpha = 0.5, colour = "darkgreen")+
geom_hline(yintercept = 50, colour = "red", linetype = "dashed")+
geom_smooth(method = "lm", colour = "blue")+
theme_gray()
````

Here we can see our initial results showing some sort of relationship, **it is positive (that is as the x value increases so does the y value)**, x = passer_rating and y= win_percentage for our Quarterbacks.

```{r Correlation Coefficient1}
#Check if there is a correlation
cor (x = df2$passer_rating, y = df2$win_percentage, method = "pearson")
````

As suggested the correlation is positive, but it is not strong as it is not that close to 1.

## Simple Linear Regression

```` {r Linear Regression}
#Check the data relating a a simple linear regression
fit <- lm(win_percentage ~ passer_rating, data = df2)

summary(fit)

tidy(fit, conf.int = TRUE)
````

The intercept, which is a measure of Y, when X = 0. The data tells us that **a Quarterback with a passer_rating of 0, should on average have a win_percentage of -50.730711. This is not possible as a Quarterback cannot have a win_percentage of less than 0.**

If we look at the confidence intervals, the best case scenario for a Quarterback with a passer_rating of 0 is still -5.393196, worst case scenario is -96.068227

The slope, is a measure of how much Y changes if we increase X by 1. In this case **if we can increase our passer_rating by 1, then we will increase our win_percentage by 1.124823.**

We can look at the confidence intervals using our equation for predicted values:
Y = Intercept + (passer_rating * Slope)
y = b + mx

Quarterback with a passer_rating of 100

```` {r Equations}
100*1.605274 + -5.393196 #highest win %

100*0.644371 + -96.068227 #lowest win %

100*1.124823 + -50.73711 #probable win %
````

What these calculations show is is the **range of winning percentages for a Quarterback with a passer-rating of 110 exceed realistic probabilities.**


## Independence of Observations
````{r Durbin-Watson}
#Determine if the observations are independent
car::durbinWatsonTest(fit)
````

The Durbin-Watson statistic shows:
a value of approximately 2 (2.37), indicating that there is no correlation between the residuals and that we have independence of observations.

## Outliers
````{r Outliers}
#Check the data for outliers
std_res <- rstandard(fit)
points <- 1:length(std_res)

ggplot(data = NULL, aes(x = points, y = std_res)) +
geom_point(colour = "darkgreen") +
ylim(c(-4, 4)) +
geom_hline(yintercept = c(-3, 3), colour = "red", linetype = "dashed")
````

There do not appear to be any outliers as **all the data points are within 3 standard deviations from 0.**

## Leverage Points
````{r Leverage Points}
#Check to see if any of the potential outliers create leverage
hats <- hatvalues(fit)

ggplot(data = NULL, aes(x = points, y = hats)) +
geom_point(colour = "darkgreen")
````

There do not appear to be any leverage points as **all hatvalues are closer to 0 than 1.**

### Further investigation of the points greater than 0.125 to ensure no leverage.
````{r Leverage Points2}
hat_labels <- if_else(hats >= 0.125, paste(points), "")

ggplot(data = NULL, aes(x = points, y = hats)) +
geom_point(colour = "darkgreen") +
geom_text(aes(label = hat_labels), nudge_y = 0.01)
````
```` {r Leverage Points3}
ggplot(df2, aes(x = passer_rating, y = win_percentage))+
geom_point(alpha = 0.5, colour = "darkgreen")+
geom_hline(yintercept = 50, colour = "red", linetype = "dashed")+
geom_smooth(method = "lm", colour = "blue")+
theme_gray()+
geom_text(aes(label = hat_labels), nudge_y = 5)
````

These three points (26, 27 & 28) **do not seem to show overly high levels of leverage or influence.**

## Influential Points
````{r Cooks Distance}
#Check for influence on the linear regression by potential outliers
cook <-cooks.distance(fit)
cook_labels <- if_else(cook >= 0.075, paste(points), "")

ggplot(data = NULL, aes(x =points, y = cook)) +
geom_point(colour = "darkgreen")+
geom_text(aes(label = cook_labels), nudge_y =0.01)
````

````{r Cooks Distance2}
ggplot(df2, aes(x = passer_rating, y = win_percentage))+
geom_point(alpha = 0.5, colour = "darkgreen")+
geom_hline(yintercept = 50, colour = "red", linetype = "dashed")+
geom_smooth(method = "lm", colour = "blue")+
theme_gray()+
geom_text(aes(label = cook_labels), nudge_y = 5)
````

It would appear that are two points (4 & 11) that are potentially influencing the data, and possibly point 28.

Point 4 appears to be a low leverage & low influence point.

Point 11 can be deemed as high leverage and high influence.

Point 28 may have a moderate leverage and moderate influence.

We should re-run our analysis removing points 11 & 28. Object 28 is however case_no 29.

## Re-run Linear Regression
````{r Outliers2}
#Removal of high influence points
outliers <- c(11, 29) #Object 28 is however case_no 29.
filtered_df2 <- df2 %>%
  filter(!case_no %in% outliers)
````

```` {r Linear Regression2}
fit2 <- lm(win_percentage ~ passer_rating, data = filtered_df2)

summary(fit2)

tidy(fit2, conf.int = TRUE)
````

````{r Plot2} 
#High influence points removed and re-run our intital model
ggplot(filtered_df2, aes(x = passer_rating, y = win_percentage))+
geom_point(alpha = 0.5, colour = "darkgreen")+
geom_hline(yintercept = 50, colour = "red", linetype = "dashed")+
geom_smooth(method = "lm", colour = "blue")+
theme_light()
````
There appears to minimal change to:
slope
intercept
R-squared value

so it appears that there is no need to remove this potential High Influence data points.

## Homoscedasticity
````{r Homoscedasticity}
#Plotting of the residuals against the fitted values
res <- residuals(fit)
fitted <- predict(fit)

ggplot(df2, aes(x = fitted, y = res))+
geom_point(colour = "purple")+
geom_hline(yintercept = 0, colour = "red", linetype = "dashed")+
geom_smooth(se = FALSE)
````
There appears to be no evidence of heteroscedasticity.

### Normality
````{r Normality}
#Is there a normal distribution

ggplot(data =NULL, aes(x= res))+
  geom_histogram( colour = "white", fill = "blue", binwidth = 10)
````

````{r Q-Q Plot}
ggplot(data = NULL, aes(sample =res))+
  stat_qq() + stat_qq_line()
````

The data points appear moderately normally distributed.

## Interpretation

**Considering the original questions, which were:
Do quarterbacks with a higher passer rating have more wins?
Do quarterbacks with a lower passer rating have fewer wins?
How many wins can we expect to get when we have a starting quarterback with a particular passer rating?**

The data has a moderate correlation between Quarterback passer_rating and win_percentage. However, further exploratory analysis shows that passer_rating should not be the only metric used. 

On average if a Quarterback can improve their passer_rating 1 point, they can improve their teams win_percentage 1.12%, so an improvement of 20 passer_rating points can make a significant difference (22%) to a teams prospect of winning games and being successful.
